Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2022-12-09 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex

\author{
Hanne Oberman\\Utrecht University \And Johanna Munoz Avila\\University
Medical Center Utrecht \AND Valentijn de Jong\\University Medical
Center Utrecht \And Gerko Vink\\Utrecht University \AND Thomas
Debray\\University Medical Center Utrecht
}
\title{Imputation of Incomplete Multilevel Data with \pkg{mice}}

\Plainauthor{Hanne Oberman, Johanna Munoz Avila, Valentijn de
Jong, Gerko Vink, Thomas Debray}
\Plaintitle{Imputation of Incomplete Multilevel Data with mice}
\Shorttitle{Multilevel \pkg{mice}}

@@

\Address{
    Hanne Oberman\\
    Utrecht University\\
    Padualaan 14\\
3584 CH Utrecht\\
  E-mail: \email{h.i.oberman@uu.nl}\\
  URL: \url{https://hanneoberman.github.io/}\\~\\
          }

@@

Many datasets include individuals from multiple settings, geographic
regions, or even different studies. In the simplest case, individuals
(e.g., students) are nested within so-called clusters (e.g., school
classes). More complex clustered structures may occur when there are
multiple hierarchical levels (e.g., patients within hospitals within
regions or countries), or when the clustering is non-nested (e.g.,
electronic health record data from diverse settings and populations
within large databases). In general, individuals from the same cluster
tend to be more similar than individuals from other clusters. In
statistical terms, this implies that observations from the same cluster
are correlated. If this correlation is left unaddressed, estimates of
\emph{p} values, confidence intervals even model parameters are prone to
bias \citep{loca01}. {[}TODO: make a link to imputation methods, which
require adequate handling and propagation of variance; we are not
recommending the adoption of multilevel models for data analysis here,
but rather for imputation.{]} Statistical methods for clustered data
typically adopt hierarchical models that explicitly describe the
grouping of observations. These models are also know as `multilevel
models', `hierarchical models', `mixed effect models' and `random effect
models'. Table \ref{tab:clus} provides an overview of some key concepts
in multilevel modeling.

@@

Imputation of missing data requires to consider the mechanism behind the
missingness. Rubin proposed to distinguish between data that are missing
completely at random (MCAR), data that are missing at random (MAR) and
data that are missing not at random (MNAR; see Table \ref{tab:miss}).
For each of these three missingness generating mechanisms, different
imputation strategies are warranted \citet{yuce08} and \citet{hox15}. We
here consider the general case that data are MAR, and expand on special
MNAR situations.

@@

The \proglang{R} package \pkg{mice} has become the de-facto standard for
imputation by chained equations, which iteratively solves the
missingness on a variable-by-variable basis. \pkg{mice} is known to
yield valid inferences under many different missing data circumstances
\citep{buur18}. However, commonly used imputation methods were not
designed for use in clustered data and usually generate observations
that are independent. For this reason, we discuss how the \proglang{R}
package \pkg{mice} can be used to impute multilevel data.

@@

This papers serves as a tutorial for imputing incomplete multilevel data
with \pkg{mice} in \proglang{R}. We provide practical guidelines and
code snippets for different missing data situations, including
non-ignorable mechanisms. For reasons of brevity, we focus on multilevel
imputation by chained equations with \pkg{mice} exclusively; other
imputation methods and packages (e.g., \pkg{jomo} and \pkg{mdmb}) are
outside the scope of this tutorial. Assumed knowledge includes basic
familiarity with multilevel imputation \citep[see e.g.][ and
\citet{grun18}]{audi18} and the \pkg{lme4} notation for multilevel
models (see Table \ref{tab:mod}).

@@

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     4.932     0.274    18.012     4.420     0.000    19.514     0.964 
sex             0.859     0.284     3.026     4.216     0.036    37.601     0.981

                            Estimate 
Intercept~~Intercept|school    0.265 
Residual~~Residual             1.007 
ICC|school                     0.213

@@

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     5.054     0.266    19.029     4.554     0.000    14.936     0.954 
sex             0.732     0.306     2.393     4.233     0.071    34.886     0.980

                            Estimate 
Intercept~~Intercept|school    0.318 
Residual~~Residual             1.257 
ICC|school                     0.208

@@

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     4.714     0.193    24.432     5.032     0.000     8.224     0.919 
sex             1.286     0.288     4.460     4.218     0.010    37.215     0.981

                            Estimate 
Intercept~~Intercept|school    0.299 
Residual~~Residual             1.050 
ICC|school                     0.228

@@

\begin{CodeChunk}
\begin{CodeInput}
R> fit <- glmer(mod, family = "binomial", data = na.omit(dat)) # fit the model
R> tidy(fit, conf.int = TRUE, exponentiate = TRUE)             # print estimates
\end{CodeInput}
\begin{CodeOutput}
# A tibble: 11 x 9
   effect   group term        estim~1 std.er~2 stati~3   p.value conf.~4 conf.~5
   <chr>    <chr> <chr>         <dbl>    <dbl>   <dbl>     <dbl>   <dbl>   <dbl>
 1 fixed    <NA>  (Intercept)  0.0863  0.0182   -11.6   2.99e-31  0.0571   0.130
 2 fixed    <NA>  typeRCT      0.757   0.137     -1.54  1.22e- 1  0.531    1.08 
 3 fixed    <NA>  age          1.03    0.00265   12.9   7.40e-38  1.03     1.04 
 4 fixed    <NA>  motor_scor~  0.651   0.0732    -3.82  1.34e- 4  0.522    0.811
 5 fixed    <NA>  motor_scor~  0.489   0.0555    -6.30  2.97e-10  0.391    0.611
 6 fixed    <NA>  motor_scor~  0.274   0.0321   -11.0   2.28e-28  0.218    0.345
 7 fixed    <NA>  pupilNone    3.20    0.317     11.7   8.18e-32  2.63     3.88 
 8 fixed    <NA>  pupilOne     1.75    0.195      5.06  4.27e- 7  1.41     2.18 
 9 fixed    <NA>  ctIII        2.41    0.268      7.89  3.05e-15  1.94     2.99 
10 fixed    <NA>  ctIV/V       2.30    0.214      8.95  3.55e-19  1.92     2.76 
11 ran_pars name  sd__(Inter~  0.230  NA         NA    NA        NA       NA    
# ... with abbreviated variable names 1: estimate, 2: std.error, 3: statistic,
#   4: conf.low, 5: conf.high
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> fit <- imp %>% 
+   with(glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial")) 
R> tidy(pool(fit))
\end{CodeInput}
\begin{CodeOutput}
                     term    estimate   std.error  statistic      p.value
1             (Intercept) -2.33813790 0.340459131  -6.867602 6.978862e-12
2                    type -0.41509929 0.179936773  -2.306918 2.107792e-02
3                     age  0.03034813 0.001561834  19.431086 0.000000e+00
4 as.factor(motor_score)2 -0.66052409 0.068846376  -9.594174 0.000000e+00
5 as.factor(motor_score)3 -1.05030912 0.070059114 -14.991756 0.000000e+00
6 as.factor(motor_score)4 -1.49637069 0.071492961 -20.930322 0.000000e+00
7                   pupil  0.48922641 0.038536483  12.695149 0.000000e+00
8                      ct  0.42634415 0.028388807  15.018037 0.000000e+00
             b          df dfcom          fmi       lambda m          riv
1 9.395965e-04  8668.10341 11013 0.0099557149 0.0097273074 5 0.0098228573
2 1.729929e-05 10991.51028 11013 0.0008229571 0.0006411648 5 0.0006415761
3 8.332782e-09 10482.95010 11013 0.0042891736 0.0040992241 5 0.0041160969
4 5.649577e-05  6979.24803 11013 0.0145856016 0.0143032580 5 0.0145108098
5 3.003051e-05  9526.87220 11013 0.0075503299 0.0073420043 5 0.0073963080
6 3.329522e-05  9362.42789 11013 0.0080288174 0.0078169354 5 0.0078785213
7 3.337070e-04    54.63795 11013 0.2949938851 0.2696512493 5 0.3692088868
8 1.180517e-05  5893.23780 11013 0.0179108151 0.0175775781 5 0.0178920775
          ubar
1 1.147849e-01
2 3.235648e-02
3 2.429325e-06
4 4.672029e-03
5 4.872243e-03
6 5.071289e-03
7 1.084612e-03
8 7.917581e-04
\end{CodeOutput}
\begin{CodeInput}
R> as.mitml.result(fit)
\end{CodeInput}
\begin{CodeOutput}
[[1]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10522.944 10588.713 -5252.472 10504.944     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2913  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.32713                 -0.41953                  0.03031  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.65651                 -1.04719                 -1.49327  
                  pupil                       ct  
                0.48990                  0.42335  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

[[2]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10525.723 10591.492 -5253.861 10507.723     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2937  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.30150                 -0.41940                  0.03024  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.65497                 -1.04451                 -1.49454  
                  pupil                       ct  
                0.47281                  0.42888  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

[[3]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
     AIC      BIC   logLik deviance df.resid 
10512.28 10578.05 -5247.14 10494.28    11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2852  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.35320                 -0.41028                  0.03034  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.66933                 -1.05474                 -1.50379  
                  pupil                       ct  
                0.49491                  0.42442  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

[[4]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
     AIC      BIC   logLik deviance df.resid 
10530.02 10595.79 -5256.01 10512.02    11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2923  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.32670                 -0.41286                  0.03037  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.66801                 -1.04770                 -1.50070  
                  pupil                       ct  
                0.47210                  0.43109  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

[[5]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10488.615 10554.383 -5235.307 10470.615     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2939  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.38215                 -0.41343                  0.03049  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.65379                 -1.05740                 -1.48956  
                  pupil                       ct  
                0.51641                  0.42397  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

@@

\begin{CodeChunk}
\begin{CodeInput}
R> imp <- mice( hivdata, # dataset with missing values
+                         m = 10,   # number of imputations
+                         maxit = 1,
+                         seed = 1234, #seed attached to the dataID
+                         meth = meth, #imputation method vector
+                         pred = pred, #imputation predictors matrix
+                         print = T,
+                         meta_method="reml",
+                         pmm=FALSE)
\end{CodeInput}
\begin{CodeOutput}

 iter imp variable
  1   1  hiv
  1   2  hiv
  1   3  hiv
  1   4  hiv
  1   5  hiv
  1   6  hiv
  1   7  hiv
  1   8  hiv
  1   9  hiv
  1   10  hiv
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{itemize}
\item
  Adding some kind of help function to mice that suggests a suitable
  predictor matrix to the user, given a certain analysis model.
\item
  Adding a \texttt{multilevel\_ampute()} wrapper function in mice.
\item
  Exporting \texttt{mids} objects to other packages like \texttt{lme4}
  or \texttt{coxme}?
\item
  Adding a ICC=0 dataset to show that even if there is no clustering it
  doesn't hurt.
\item
  Show use case for deductive imputation for cluster level variables?
\item
  env dump in repo
\end{itemize}

>>>>

\author{
Hanne I. Oberman\\Utrecht University \And Johanna Munoz Avila\\Julius
Center for Health Sciences and Primary Care, University Medical Center
Utrecht, Utrecht University, Utrecht, The Netherlands \And Thomas P. A.
Debray\\Julius Center for Health Sciences and Primary Care, University
Medical Center Utrecht, Utrecht University, Utrecht, The
Netherlands \And Gerko Vink\\Utrecht University \And Valentijn M. T. de
Jong\\Julius Center for Health Sciences and Primary Care, University
Medical Center Utrecht, Utrecht University, Utrecht, The Netherlands
\AND Data Analytics and Methods Task Force, European Medicines Agency,
Amsterdam, The Netherlands
}
\title{Imputation of Incomplete Multilevel Data with \pkg{mice}}

\Plainauthor{Hanne I. Oberman, Johanna Munoz Avila, Thomas P. A.
Debray, Gerko Vink, Valentijn M. T. de Jong}
\Plaintitle{Imputation of Incomplete Multilevel Data with mice}
\Shorttitle{Multilevel \pkg{mice}}

@@

\Address{
    Hanne I. Oberman\\
    Utrecht University\\
    Padualaan 14\\
3584 CH Utrecht\\
  E-mail: \email{h.i.oberman@uu.nl}\\
  URL: \url{https://hanneoberman.github.io/}\\~\\
          }

@@

Many datasets include individuals from multiple settings, geographic
regions, or even different studies. In the simplest case, individuals
(e.g., students) are nested within so-called clusters (e.g., school
classes). More complex clustered structures may occur when there are
multiple hierarchical levels (e.g., patients within hospitals within
regions or countries), or when the clustering is non-nested (e.g.,
electronic health record data from diverse settings and populations
within large databases). In general, individuals from the same cluster
tend to be more similar than individuals from other clusters. In
statistical terms, this implies that observations from the same cluster
are correlated. If this correlation is left unaddressed, estimates of
\emph{p} values, confidence intervals even model parameters are prone to
bias \citep{loca01}. {[}TODO: make a link to imputation methods, which
require adequate handling and propagation of variance; we are not
recommending the adoption of multilevel models for data analysis here,
but rather for imputation. VMTdJ: ``I would only make that link after
introducing missing data, and missing data is not mentioned yet, so I
would do that in the next section''{]} Statistical methods for clustered
data typically adopt hierarchical models that explicitly describe the
grouping of observations. These models are also known as `multilevel
models', `hierarchical models', `mixed effect models', `random effect
models', and in the context of time-to-event data as `frailty models'.
Table \ref{tab:clus} provides an overview of some key concepts in
multilevel modeling.

@@

Imputation of missing data requires consideration of the mechanism
behind the missingness. Rubin proposed to distinguish between data that
are missing completely at random (MCAR), data that are missing at random
(MAR) and data that are missing not at random (MNAR; see Table
\ref{tab:miss}). For each of these three missingness generating
mechanisms, different imputation strategies are warranted \citet{yuce08}
and \citet{hox15}. We here consider the general case that data are MAR,
and expand on certain MNAR situations.

@@

The \proglang{R} package \pkg{mice} has become the de-facto standard for
imputation by chained equations, which iteratively solves the
missingness on a variable-by-variable basis. \pkg{mice} is known to
yield valid inferences under many different missing data circumstances
\citep{buur18}. However, commonly used imputation methods were not
designed for use in clustered data and usually generate observations
that are independent whereas multilevel data are dependent. For this
reason, we discuss how the \proglang{R} package \pkg{mice} can be used
to impute multilevel data.

@@

This papers serves as a tutorial for imputing incomplete multilevel data
with \pkg{mice} in \proglang{R}. We provide practical guidelines and
code snippets for different missing data situations, including
non-ignorable mechanisms. For reasons of brevity, we focus on multilevel
imputation by chained equations with \pkg{mice} exclusively; other
imputation methods and packages (e.g., \pkg{jomo} and \pkg{mdmb}) are
outside the scope of this tutorial. {[}TODO: change this sentence
because we do not assume familiarity: ``Assumed knowledge includes basic
familiarity with multilevel imputation \citep[see e.g.][ and
\citet{grun18}]{audi18} and the \pkg{lme4} notation for multilevel
models (see Table \ref{tab:mod})''{]}.

@@

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     4.986     0.194    25.704     4.917     0.000     9.201     0.927 
sex             0.818     0.357     2.290     4.130     0.082    62.169     0.989

                            Estimate 
Intercept~~Intercept|school    0.272 
Residual~~Residual             0.975 
ICC|school                     0.227

@@

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     5.186     0.254    20.457     4.706     0.000    11.816     0.942 
sex             0.760     0.219     3.479     4.609     0.020    13.627     0.950

                            Estimate 
Intercept~~Intercept|school    0.345 
Residual~~Residual             1.578 
ICC|school                     0.182

@@

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     5.000     0.356    14.027     4.283     0.000    28.723     0.976 
sex             1.056     0.465     2.270     4.105     0.084    76.708     0.991

                            Estimate 
Intercept~~Intercept|school    0.294 
Residual~~Residual             1.347 
ICC|school                     0.187

@@

\begin{CodeChunk}
\begin{CodeInput}
R> fit <- glmer(mod, family = "binomial", data = na.omit(dat)) # fit the model
R> tidy(fit, conf.int = TRUE, exponentiate = TRUE)             # print estimates
\end{CodeInput}
\begin{CodeOutput}
# A tibble: 11 x 9
   effect   group term        estim~1 std.er~2 stati~3   p.value conf.~4 conf.~5
   <chr>    <chr> <chr>         <dbl>    <dbl>   <dbl>     <dbl>   <dbl>   <dbl>
 1 fixed    <NA>  (Intercept)  0.0863  0.0182   -11.6   3.00e-31  0.0571   0.130
 2 fixed    <NA>  typeRCT      0.757   0.137     -1.54  1.22e- 1  0.531    1.08 
 3 fixed    <NA>  age          1.03    0.00265   12.9   7.40e-38  1.03     1.04 
 4 fixed    <NA>  motor_scor~  0.651   0.0732    -3.82  1.34e- 4  0.522    0.811
 5 fixed    <NA>  motor_scor~  0.489   0.0555    -6.30  2.97e-10  0.391    0.611
 6 fixed    <NA>  motor_scor~  0.274   0.0321   -11.0   2.28e-28  0.218    0.345
 7 fixed    <NA>  pupilNone    3.20    0.317     11.7   8.18e-32  2.63     3.88 
 8 fixed    <NA>  pupilOne     1.75    0.195      5.06  4.27e- 7  1.41     2.18 
 9 fixed    <NA>  ctIII        2.41    0.268      7.89  3.05e-15  1.94     2.99 
10 fixed    <NA>  ctIV/V       2.30    0.214      8.95  3.56e-19  1.92     2.76 
11 ran_pars name  sd__(Inter~  0.230  NA         NA    NA        NA       NA    
# ... with abbreviated variable names 1: estimate, 2: std.error, 3: statistic,
#   4: conf.low, 5: conf.high
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> fit <- imp %>% 
+   with(glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial")) 
R> tidy(pool(fit))
\end{CodeInput}
\begin{CodeOutput}
                     term    estimate   std.error  statistic      p.value
1             (Intercept) -2.33931814 0.337744863  -6.926288 4.565459e-12
2                    type -0.41950933 0.179122441  -2.342026 1.919717e-02
3                     age  0.03042682 0.001565356  19.437637 0.000000e+00
4 as.factor(motor_score)2 -0.66613691 0.069302767  -9.611981 0.000000e+00
5 as.factor(motor_score)3 -1.04915293 0.070119780 -14.962297 0.000000e+00
6 as.factor(motor_score)4 -1.50206863 0.071907169 -20.888997 0.000000e+00
7                   pupil  0.49317740 0.035150534  14.030438 0.000000e+00
8                      ct  0.42802910 0.029616105  14.452579 0.000000e+00
             b         df dfcom          fmi       lambda m          riv
1 2.240527e-04 10819.9749 11013 0.0025413258 0.0023569693 5 0.0023625377
2 2.066282e-05 10984.4464 11013 0.0009546917 0.0007728065 5 0.0007734042
3 1.927783e-08  8774.5121 11013 0.0096665946 0.0094408908 5 0.0095308707
4 1.096200e-04  3559.8620 11013 0.0279345878 0.0273886158 5 0.0281598759
5 3.652274e-05  8968.6823 11013 0.0091347491 0.0089138126 5 0.0089939832
6 9.004098e-05  4952.3514 11013 0.0212918407 0.0208966706 5 0.0213426612
7 1.183186e-04   293.7825 11013 0.1208778877 0.1149133407 5 0.1298328695
8 6.936809e-05   425.1630 11013 0.0991319347 0.0949041246 5 0.1048553276
          ubar
1 1.138027e-01
2 3.206005e-02
3 2.427206e-06
4 4.671329e-03
5 4.872956e-03
6 5.062592e-03
7 1.093578e-03
8 7.938720e-04
\end{CodeOutput}
\begin{CodeInput}
R> as.mitml.result(fit)
\end{CodeInput}
\begin{CodeOutput}
[[1]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10509.778 10575.547 -5245.889 10491.778     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2939  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
                -2.3395                  -0.4185                   0.0304  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
                -0.6597                  -1.0504                  -1.5006  
                  pupil                       ct  
                 0.4893                   0.4302  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

[[2]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10501.337 10567.106 -5241.668 10483.337     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2896  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.35407                 -0.42416                  0.03023  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.66302                 -1.03875                 -1.50219  
                  pupil                       ct  
                0.50367                  0.43246  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

[[3]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10500.594 10566.363 -5241.297 10482.594     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2902  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.35434                 -0.41988                  0.03042  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.65515                 -1.04987                 -1.49095  
                  pupil                       ct  
                0.49221                  0.43460  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

[[4]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10520.673 10586.442 -5251.336 10502.673     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2891  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.32285                 -0.42261                  0.03062  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.67111                 -1.05287                 -1.49941  
                  pupil                       ct  
                0.50321                  0.41360  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 2 lme4 warnings

[[5]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10532.391 10598.160 -5257.196 10514.391     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2866  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.32583                 -0.41243                  0.03045  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.68169                 -1.05384                 -1.51715  
                  pupil                       ct  
                0.47746                  0.42928  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

@@

\begin{CodeChunk}
\begin{CodeInput}
R> imp <- mice( hivdata, # dataset with missing values
+                         m = 10,   # number of imputations
+                         maxit = 1,
+                         seed = 1234, #seed attached to the dataID
+                         meth = meth, #imputation method vector
+                         pred = pred, #imputation predictors matrix
+                         print = T,
+                         meta_method="reml",
+                         pmm=FALSE)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{itemize}
\item
  Adding evaluations of the imputations such as convergence checks
\item
  Adding some kind of help function to mice that suggests a suitable
  predictor matrix to the user, given a certain analysis model.
\item
  Adding a \texttt{multilevel\_ampute()} wrapper function in mice.
\item
  Exporting \texttt{mids} objects to other packages like \texttt{lme4}
  or \texttt{coxme}?
\item
  Adding a ICC=0 dataset to show that even if there is no clustering it
  doesn't hurt.
\item
  Show use case for deductive imputation for cluster level variables?
\item
  env dump in repo
\end{itemize}
